---
title: "Reading FreeSurfer format brain imaging data files with freesurferformats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading FreeSurfer format brain imaging data files with freesurferformats}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this document, we show how to read brain imaging data from [FreeSurfer](https://surfer.nmr.mgh.harvard.edu) binary files. These files are created and used by the [FreeSurfer neuroimaging software suite](https://surfer.nmr.mgh.harvard.edu) to store volume data and surface morphometry data computed from MRI brain images.


# Reading FreeSurfer data with the *freesurferdata* R package

## Background

Brain imaging data come in different formats. Typically, the data is acquired on a scanner that outputs a set of two-dimensional (2D) DICOM format images. The DICOM images are often combined into a single file that holds a 3D or 4D stack of images for further processing. Common formats include ANALYZE, NIFTI, and the MGH format used by FreeSurfer. This *freesurferdata* R package implements functions to parse data in MGH format, as well as some related FreeSurfer formats.


## Reading MGH and MGZ format files




## Reading 'curv' format files

Let's read an example morphometry data file that comes with this package. It contains vertex-wise measures of cortical thickness for the left hemisphere of a single subject in native space.

```{r}
    curvfile = system.file("extdata", "lh.thickness", package = "freesurferformats", mustWork = TRUE);
    ct = read.fs.curv(curvfile);
```

We can now have a closer look at the data and maybe plot a histogram:

```{r}
    cat(sprintf("Read data for %d vertices. Values: min=%f, mean=%f, max=%f.\n",  length(ct), min(ct), mean(ct), max(ct)));
    hist(ct, main="Histogram of cortical thickness for left hemisphere", xlab="Cortical thickness [mm]", ylab="Vertex count");
```  


## References
* See [the FreeSurfer wiki](https://surfer.nmr.mgh.harvard.edu/fswiki/FsTutorial/MghFormat) for details on the MGH format


## Alternatives and similar Packages
* The [freesurfer package by John Muschelli]( https://CRAN.R-project.org/package=freesurfer) implements an R wrapper around some FreeSurfer command line utilities. This includes functions to read MGH/MGZ files by converting them to NIFTI format using command line utilities that come with FreeSurfer, then loading the resulting NIFTI file with the [oro.nifti package](https://CRAN.R-project.org/package=oro.nifti). As such, the package requires that you have FreeSurfer installed.
* Several packages exist that read files in NIFTI format, including [oro.nifti](https://CRAN.R-project.org/package=oro.nifti) and [RNifti](https://CRAN.R-project.org/package=RNifti)
