---
title: "Writing FreeSurfer neuroimaging data with freesurferformats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing FreeSurfer neuroimaging data with freesurferformats}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this document, we show how to write brain imaging data to [FreeSurfer](https://surfer.nmr.mgh.harvard.edu) binary files.


# Writing FreeSurfer neuroimaging data with freesurferformats




## Writing 1D morphometry data or other per-vertex information (in MGH, MGZ or curv format)


Morphometry data, or *vertex-wise measures*, are data that usually describe a measure like cortical thickness or surface area over the cortex. There is one scalar value per vertex of the brain surface mesh. Of course, you could write whatever you want (p-values, effect sizes at the vertex, ...), as long as the data is scalar.

The package provides the `write.fs.morph` function to write any scalar data that does not require metadata like MR acquisition parameters or transforms. With this function, the format gets determined automatically from the file name. In the following example, we load the area and thickness values for a subject and write the product of area and thickness (which is *not* cortical volume, by the way) to new files in MGH, MGZ and curv format.

Let's first load the data:


```{r, eval = FALSE}
    library("freesurferformats");
    area = read.fs.morph(system.file("extdata", "lh.thickness", package = "freesurferformats", mustWork = TRUE));
    thickness = read.fs.morph(system.file("extdata", "lh.area.gz", package = "freesurferformats", mustWork = TRUE));
    mymorphdata = area * thickness;
```  

Now we could write our derived data like this:

```{r, eval = FALSE}
    format1 = write.fs.morph(tempfile(fileext = "mgz"), mymorphdata);
    format2 = write.fs.morph(tempfile(fileext = "mgh"), mymorphdata);
    format3 = write.fs.morph(tempfile(fileext = "curv"), mymorphdata);
```  



## Writing MGH and MGZ format files with full control over the header


When writing 3D or 4D images, the voxels alone are sometimes not enough: depending on what you intend to do with the data later on, you may want to save metadata like MR acquisition parameters or vox2ras transformation matrices in the file header. This is possible with the `write.fs.mgh` function, that gives you full control over the MGH header. The two relevant pieces of header data are:

* the `vox2ras_matrix`: a 4x4 double matrix that encodes the affine transformation from voxel indices to *x, y, z* coordinates in space
* the `mr_params`: a double vector of length 4 that contains the following values (in this order):
    + tr: repetition time in ms
    + flipangle: flip angle in radians
    + te: echo time in ms
    + ti: inversion time in ms

Here is an example that writes a file in MGH format including custom header data:

```{r, eval = FALSE}
    mgh_outfile = "mystudy/subject1/mri/shifted_brain.mgz"
    data = array(data=rep(1L, 256*256*256), dim=c(256,256,256)); # not exactly a brain, but will do.
    mr_params = c(2300, 0.1, 2., 900.)
    vox2ras_matrix = matrix(c(-1,0,0,0,  0,0,-1,0,  0,1,0,0,  127.5,-98.6273,79.0953,1.000), nrow=4, byrow = FALSE)
    write.fs.mgh(mgh_outfile, data, vox2ras_matrix=vox2ras_matrix, mr_params=mr_params);
```


Note that *if and only if* you provide a `ras2vox_matrix`, the *ras_good* flag will be set to `TRUE` in the file header. If you do not provide `mr_params`, they default to all zero.

### MGH data types

Data can be stored in the following data types in MGH files:

* 'MRI_UCHAR' : 1 byte unsigned integer
* 'MRI_SHORT' : 2 byte signed integer
* 'MRI_INT'   : 4 byte signed integer
* 'MRI_FLOAT' : 4 byte signed floating point

One can control the data type that is used through the `dtype` parameter. The default value, 'auto', will determine the MRI data type from the R data type of the `data` parameter. The following rules are applied for R data types:

* *logical* : the data gets coerced to integer, it is written as 'MRI_UCHAR'
* *integer* : written as 'MRI_INT' (no matter the range of values)
* *double*  : written as 'MRI_FLOAT'

You can manually set the 'dtype' parameter to force a certain MRI data type. E.g., you could force 'MRI_SHORT' for your integer data if you are sure that the supported value range of that data type is enough for the values contained in your data. Be aware though that chosing an unsuitable MRI data type for your data will lead to an MGH file with incorrect data.

*Hint*: If you need to save data in a certain data type, cast your data before passing them to the `write.fs.mgh` function. E.g., if you have integers but want to store them as 'MRI_FLOAT', do the following:

```{r, eval = FALSE}
    some_surface_mask = rep(1L, 163842);    
    some_surface_mask[30000:45000] = 0L;
    write.fs.mgh("regionmask_stored_as_MRI_FLOAT.mgh", as.double(some_surface_mask));
```

## Writing 'curv' format files

You can use `write.fs.curv` to write arbitrary data in binary curv format. The result is identical to using `write.fs.morph` with any filename that does not end in `mgh` or `mgz` instead.)

```{r, eval = FALSE}
    data = rnorm(120000, 2.0, 1.0);
    curvfile = "mystudy/subject1/surf/lh.random"
    write.fs.curv(curvfile, data);
```


It's worth knowing the if your filename ends with `.gz`, the file will be written in gzip format.


## Writing surface format files

You can use `write.fs.surface` to write triangular meshes in binary surface format (the format used for files like 'surf/lh.white' or 'surf/rh.pial'). A mesh is defined by a list of vertices and a list of faces.

```{r, eval = FALSE}
    vertices = matrix(rep(0.3, 15), nrow=3);     # 5 vertices
    faces = matrix(c(1L,2L,3L,2L,4L,3L,4L,5L,3L), nrow=3, byrow = TRUE);   # 3 faces

    write.fs.surface(tempfile(fileext="white"), vertices, faces);
```

The vertex indices used to define the faces should be 1-based, as used in R. They will be written 0-based to the file.


## Writing label files

Labels can be written with the `write.fs.label` function. A label is nothing but a list of vertex indices.

```{r, eval = FALSE}
output_file = tempfile();

# generate data
vertex_indices = seq(from = 10000, to=20000);

# write label to file
write.fs.label(output_file, vertex_indices);
```



## Writing color lookup table (LUT) files in ASCII format

The following example uses the `write.fs.colortable` function to write a colortable:

```{r, eval = FALSE}
colortable_df = data.frame("struct_index"=c(0, 1), "struct_name"=c("struct1", "struct2"),
                  "r"=c(80, 100), "g"=c(50, 40), "b"=c(250, 200), "a"=c(0, 0), stringsAsFactors = FALSE);

output_file = tempfile(fileext = ".txt");
write.fs.colortable(output_file, colortable_df);
```


