---
title: "Reading FreeSurfer neuroimaging data with freesurferformats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading FreeSurfer neuroimaging data with freesurferformats}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this document, we show how to read brain imaging data from [FreeSurfer](https://surfer.nmr.mgh.harvard.edu) binary files. These files are created and used by the [FreeSurfer neuroimaging software suite](https://surfer.nmr.mgh.harvard.edu) to store volume data and surface morphometry data computed from MRI brain images.


# Reading FreeSurfer neuroimaging data with freesurferformats

## Background

Brain imaging data come in different formats. Typically, the data is acquired on a scanner that outputs a set of two-dimensional (2D) DICOM format images. The 2D images are often combined into a single file that holds a 3D or 4D stack of images for further processing. Common formats include ANALYZE, NIFTI, and the MGH format used by FreeSurfer.

This *freesurferdata* R package implements functions to parse data in MGH format, as well as some related FreeSurfer formats. MGH stands for Massachusetts General Hospital, and is a binary format. The MGZ format is a compressed version of the MGH format.

**Note:** To learn how to *write* neuroimaging data with this package, read the vignette *Writing FreeSurfer neuroimaging data with freesurferformats* that comes with this package.


## Reading MGH and MGZ format files

Here is a first example for reading MGH data.


```{r}
    library("freesurferformats")
    mgh_file = system.file("extdata", "brain.mgz", package = "freesurferformats", mustWork = TRUE)
    brain = read.fs.mgh(mgh_file)
    cat(sprintf("Read voxel data with dimensions %s. Values: min=%d, mean=%f, max=%d.\n", paste(dim(brain), collapse = 'x'), min(brain), mean(brain), max(brain)));
```

Now, `brain` is an *n*-dimensional matrix, where *n* depends on the data in the MGZ file. A conformed FreeSurfer volume like `brain.mgz` typically has 4 dimensions and 256 x 256 x 256 x 1 = 16777216 voxels. The final dimension, which is 1 here, means it has only a single time point or *frame*. In this case, the file was compressed and in MGZ format, but the function does not care, it works for both MGH and MGZ.

If you need the header data, read on.

### Accessing metadata from the MGH header

To access not only the volume data, but also the header, call `read.fs.mgh` like this:

```{r}
    brain_with_hdr = read.fs.mgh(mgh_file, with_header = TRUE);
    brain = brain_with_hdr$data;   # as seen before, this is what we got in the last example.
    header = brain_with_hdr$header;   # as seen before, this is what we got in the last example.
```


Now you have acces to the following header data:

```{r, eval = FALSE}
    header$dtype           # int, one of: 0=MRI_UCHAR; 1=MRI_INT; 3=MRI_FLOAT; 4=MRI_SHORT
    header$ras_good_flag   # int, 0 or 1. Whether the file contains a valid vox2ras matrix and ras_xform (see header$vox2ras_matrix below)
    header$has_mr_params   # int, 0 or 1. Whether the file contains mr_params (see header$mr_params below)
    header$voldim          # integer vector or length 4. The volume (=data) dimensions. E.g., c(256, 256, 256, 1) for 3D data.

```


If your MGH/MGZ file contains valid information on the vox2ras matrix and/or acquisition parameters (`mr_params`), you can access them like this for the mr params:

```{r}
    if(header$has_mr_params) {
        mr_params = header$mr_params;  
        cat(sprintf("MR acquisition parameters: TR [ms]=%f, filp angle [radians]=%f, TE [ms]=%f, TI [ms]=%f\n", mr_params[1], mr_params[2], mr_params[3], mr_params[4]));
    }
```

And like this for the `vox2ras_matrix`:

```{r}
    if(header$ras_good_flag) {
        print(header$vox2ras_matrix);
    }
```


And finally the `ras_xform`:

```{r}
    if(header$ras_good_flag) {
        print(header$ras_xform);
    }
```


### A second MGH example

The MGH/MGZ format is also used to store morphometry data mapped to standard space (fsaverage). In the following example, we read cortical thickness data in standard space, smoothed with a FWHM 25 kernel:


```{r, eval = FALSE}
    mgh_file = system.file("mystudy", "subject1", "surf", "lh.thickness.fwhm25.fsaverage.mgh")
    cortical_thickness_standard = read.fs.mgh(mgh_file)
```

Now, `cortical_thickness_standard` is a vector of *n* float values, where *n* is the number of vertices of the *fsaverage* subject's left hemisphere surface (i.e., 163842 in FreeSurfer 6).

### Hint: Further image processing and visualization for volumes

If all you need is to perform statistical analysis of the data in the MGH file, you are ready to do that after loading. If you need access to more image operations, I would recommend to convert the data to a NIFTI object. E.g., if you have [oro.nifti](https://CRAN.R-project.org/package=oro.nifti ) installed, you could visualize the `brain` data we loaded earlier like this:

```{r, eval = FALSE}
    oro.nifti::orthographic(oro.nifti::nifti(brain))
```


## Reading 'curv' format files

Let's read an example morphometry data file that comes with this package. It contains vertex-wise measures of cortical thickness for the left hemisphere of a single subject in native space.

```{r}
    library("freesurferformats")
    curvfile = system.file("extdata", "lh.thickness", package = "freesurferformats", mustWork = TRUE)
    ct = read.fs.curv(curvfile)
```

Now, `ct` is a vector of *n* float values, where *n* is the number of vertices of the surface mesh the data belongs to (usually `surf/lh.white`). The number of vertices differs between subjects, as this is native space data.

We can now have a closer look at the data and maybe plot a histogram of cortical thickness for this subject:

```{r}
    cat(sprintf("Read data for %d vertices. Values: min=%f, mean=%f, max=%f.\n",  length(ct), min(ct), mean(ct), max(ct)))
    hist(ct, main="lh cortical thickness", xlab="Cortical thickness [mm]", ylab="Vertex count")
```  


## Reading morphometry data, no matter the format

The package provides a wrapper function to read morphometry data, no matter the format. It always returns data as a vector and automatically determines the format from the file name. Here we use the function to read the file from the last example:


```{r}
    morphfile1 = system.file("extdata", "lh.thickness", package = "freesurferformats", mustWork = TRUE)
    thickness_native = read.fs.morph(morphfile1)
```  


And here is an example for an MGZ file:

```{r}
    morphfile2 = system.file("extdata", "lh.curv.fwhm10.fsaverage.mgz", package = "freesurferformats", mustWork = TRUE)
    curv_standard = read.fs.morph(morphfile2)
    hist(curv_standard, main="lh std curvature", xlab="Mean Curvature [mm^-1]", ylab="Vertex count", xlim=c(-0.1, 0.1))
```  



## Reading annotation files

An annotation file contains a cortical parcellation for a subject, based on a brain atlas. It contains a label for each vertex of a surface, and that label assigns this vertex to one of a set of atlas regions. The file format also contains a colortable, which assigns a color code to each atlas region. An example file would be `labels/lh.aparc.annot` for the `aparc` (Desikan) atlas.

Let's read an example annotation file that comes with this package:

```{r}
    library("freesurferformats")
    annotfile = system.file("extdata", "lh.aparc.annot.gz", package = "freesurferformats", mustWork = TRUE)
    annot = read.fs.annot(annotfile)
```


**Note:** The example file that comes with this package was gzipped to save space. While this is not typical for annot files, the *read.fs.annot* function handles it automatically if the filename ends with *.gz*.


## Working with annotation data

As mentioned earlier, such a file contains various pieces of information. Let us investigate the labels and the atlas region names for some vertices first:

```{r}    
    num_vertices_total = length(annot$vertices);
    for (vert_idx in c(1, 5000, 123456)) {
        cat(sprintf("Vertex #%d with zero-based index %d has label code '%d' which stands for atlas region '%s'\n", vert_idx, annot$vertices[vert_idx], annot$label_codes[vert_idx], annot$label_names[vert_idx]));
    }
```

Now, we will focus on the colortable. We will list the available regions and their color codes.

```{r}
    ctable = annot$colortable$table;
    regions = annot$colortable$struct_names;
    for (region_idx in seq_len(annot$colortable$num_entries)) {
        cat(sprintf("Region #%d called '%s' has RGBA color (%d %d %d %d) and code '%d'.\n", region_idx, regions[region_idx], ctable[region_idx,1], ctable[region_idx,2], ctable[region_idx,3], ctable[region_idx,4], ctable[region_idx,5]));
    }
```

That's all the information you can get from an annotation file.



## References
* See [the FreeSurfer wiki](https://surfer.nmr.mgh.harvard.edu/fswiki/FsTutorial/MghFormat) for details on the MGH format


## Alternatives, similar and related Packages
* The [freesurfer package by John Muschelli]( https://CRAN.R-project.org/package=freesurfer) implements an R wrapper around some FreeSurfer command line utilities. This includes functions to read MGH/MGZ files by converting them to NIFTI format using command line utilities that come with FreeSurfer, then loading the resulting NIFTI file with the [oro.nifti package](https://CRAN.R-project.org/package=oro.nifti). As such, the package requires that you have FreeSurfer installed.
* Several packages exist that read files in NIFTI format, including [oro.nifti](https://CRAN.R-project.org/package=oro.nifti) and [RNifti](https://CRAN.R-project.org/package=RNifti)
